---
title: "R basics"
author: "Jonas Schöley"
date: "September 5th, 2017"
output:
  github_document:
    toc: yes
  pdf_document:
    toc: yes
---

Todays concepts
---------------

  - arithmetic calculations
  - order of evaluation
  - objects
  - assignments
  - the workspace
  - indexing
  - vectors
  - vectorization
  - data frames
  - functions
  - arguments
  - logical comparisons

Todays operators
----------------

  - `+`, `-`, `/`, `*`, `^`
  - `[]`
  - `<-`
  - `>=`
  - `:`

Todays functions
----------------

  - `data.frame()`
  - `plot()`, `lines()`
  - `log()`, `exp()`
  - `function()`
  - `uniroot()`

When will 10 billion people live on earth?
------------------------------------------

In 2017 around 7.5 billion live on planet earth. Ten years earlier the number was 6.6 billion. We feed this information into R by *assigning* it to an *object* via the *assignment operator* `<-`.

```{r}
N2017 <- 7.5e9
N2007 <- 6.6e9
```

We choose the names `N2017` and `N2007` for our objects. It is a good idea to pick a name which lets you remember the content of the object. Some mild restrictions apply to naming:

  > "A syntactically valid name consists of letters, numbers and the dot or
  > underline characters and starts with a letter or the dot not followed by a
  > number. Names such as '.2way' are not valid, and neither are the reserved
  > words." (from `?make.names`)

### Arithmetic growth

Now that we know the world population at two points in time we can calculate the average absolute annual growth rate. Doing basic arithmetic in R feels much like using a desk calculator.

```{r}
# an average annual growth of 90 million people
R <- (N2017-N2007)/10
```

Again, we assign the result of our calculation to a new object named `R`, the absolute annual growth rate.

We use absolute growth rates if we want to model linear growth, i.e. a situation where the population number $N$ is a linear function of time:

$$
N(t+n) = N(t) + nR
$$

So our projections for the population numbers in 2018 and 2019 are

```{r}
N2017+R # 2018
N2017+R*2 # 2019
```

What about the next 10 years? It would be tedious to type it out manually so we make use of Rs vector arithmetic.

```{r}
# next 10 years
N2017+R*1:10
```

Using the `:` operator we have constructed an integer sequence from 1 to 10, i.e. a vector object. Multiplying a *scalar* (a vector with a single element) like `R` with a vector like `1:10` produces a vector with entries $R\cdot 1, R\cdot 2, \ldots, R \cdot 10$. Vector (and matrix) operations are at the heart of R. Usually we don't need to state what we want on an element-by-element basis but instead work with complete vectors. Writing R code in a way that makes use of the vector and matrix features in R is called *vectorization* -- an important concept for writing efficient R code.

We want to get an estimate of when the world population will hit 10 million people. So we will linearly project 50 years into the future and store our results in a `data.frame` -- think of it as a table.

```{r}
arith_growth <-
  data.frame(
    year = 2017 + 1:50,
    N = N2017 + R*1:50
  )
arith_growth
```

Our data frame has two columns and 50 rows. The columns have names `year` and `N`. The advantage of using a data frame to store your calculation results is the same advantage as using a table: all your results in a single place. Note that in order to "see" (`print` in R lingo) the table we have to explicitly call the object we stored the table in (`arith_growth`).

The function `plot()`, if applied to a data frame with two numeric columns, plots the first column against the second.

```{r}
plot(arith_growth)
```

```{r}
# all rows with a value of N larger or equal to 10 billion
arith_growth[arith_growth$N >= 10e9,]
# all years of rows with a value of N larger or equal to 10 billion
arith_growth[arith_growth$N >= 10e9, 'year'][1]
# the first year where N gets larger or equal to 10 billion
arith_growth[arith_growth$N >= 10e9, 'year'][1]
```

Geometric growth
----------------

$$
N(t+n) = N(t) \cdot (1+r)^n
$$

```{r}
# an average annual growth rate of ~1.3%
r <- (1+(N2017-N2007)/N2007)^(1/10) - 1

# next 10 years
N2007*(1+r)^(1:10) # what's with that: N0*(1+r)^1:10

geom_growth <-
  data.frame(
    year = 2017 + 1:100,
    N = N2017*(1+r)^(1:100)
  )
geom_growth
plot(arith_growth)
lines(geom_growth, type = 'p')
geom_growth[geom_growth$N >= 10e9, 'year'][1]
```

Exponential growth
------------------

$$
N(t+n) = N(t) \cdot \exp(rn)
$$

```{r}
r = log(N2017/N2007)/10

N2007*exp(r*1:10)

ExpGrowth <- function (x, N0, r) {
  N0*exp(r*x)
}

years_until_10b <- uniroot(function (x) {ExpGrowth(x, N2007, r)-10e9}, interval = c(0, 100))
years_until_10b
```

Further reading
---------------

- Riffe etal. (2017). EDSD Computer Programming. Chapter 1. Basic R.
- Smith, D. P., & Keyfitz, N. (2013). Mathematical Demography. Chapter 1.1.
  Berlin, Heidelberg: Springer. https://doi.org/10.1007/978-3-642-35858-6